---
title: "TCGA Explorer"
author: "Kyle Dunlap"
date: "2024-02-16"
output: github_document
---

## Introduction

This repository is used to store information about TCGA datasets as well as code on how to mine through them.
In this README, I will share with you several functions that will hopefully help you mine through TCGA data to get some useful results.
These functions use a dataset compiled from the PANCANCER subset in Xenabrowser. If you want to see how I cleaned and put this dataset together, please check out the code in this repository. This pre-cleaned dataset is stored on the University of Utah Biochemistry Fileserver. It is comprised of 12,724 patients, 20,532 genes, and 2 columns at the beginning indicating the TCGA cancer code (https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations) and the abbreviated patient ID. Below is the code used to very quickly do a final clean on this sheet, and what the output looks like.

```{r load in data, cache = TRUE, message = FALSE}
library(tidyverse)
load("all.RData")

```

```{r make head, echo = FALSE, message = FALSE}
library(tidyverse)
all <- all2
head(all)
```
This dataset is named "all" and will be the main input for all of the following functions.


# Function 1 - scatter() - Makes a scatterplot of two genes.

To use this function, you input two genes and then the cancer type (ACC, BLCA, LIHC) for example. If you want to make a large plot utilizing all 36 cancer types, you will plug in "TCGA" as the third argument.

``` {r function 1, cache = TRUE}
scatter <- function(gene1, gene2, code) {
  
  gene1 <- ensym(gene1)
  gene2 <- ensym(gene2)
  code <- ensym(code)

  filtered_data <- all |>
    select({{gene1}}, {{gene2}}, Type, sample_type) |>
    filter(if (code != "TCGA") Type == code else TRUE) |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic")
  
  cor_value <- cor(filtered_data[[gene1]], filtered_data[[gene2]])
  
  ggplot(filtered_data, aes(x = {{gene1}}, y = {{gene2}})) + 
    geom_point(alpha = 0.2) +
    annotate("text", x = min(filtered_data[[gene1]]), y = max(filtered_data[[gene2]]),
             label = paste(code, "Correlation Coefficient: r =", round(cor_value, 3)),
             hjust = 0, vjust = 0, size = 5.5) +
    geom_smooth(method = "lm") +
    theme_classic() +
    theme(axis.text.x = element_text(size = 15)) +
    theme(axis.text.y = element_text(size = 15)) +
    theme(axis.title.x = element_text(size = 15)) +
    theme(axis.title.y = element_text(size = 15)) 
}
```
Running this function gives the output below.

```{r function 1 result, message = FALSE, dpi = 300}
scatter(PHGDH, PSAT1, TCGA)
```

# Function 2 - scatter_facet() - Makes scatter plots of 2 genes for all 36 cancer types.
This function takes advantage of the facet_wrap() function in ggplot2. You will input two genes, and get 36 "mini-scatter plots", one for each cancer type. It will return both a table and a plot. To save space, I only show a tiny snippet of the plot, but you can see the whole thing by changing print(n = 3) to print(n = Inf).

```{r function 2}
scatter_facet <- function(gene1, gene2) {
  gene1 <- ensym(gene1)
  gene2 <- ensym(gene2)
  
  
  library(ggpubr)
  filtered_data <- all |>
    select({{gene1}}, {{gene2}}, Type, sample_type) |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic")
  
  cor_value <- filtered_data |>
    group_by(Type) |>
    summarize(
      cor = cor({{gene1}}, {{gene2}})
    )
  
  plot <- ggplot(filtered_data, aes(x = {{gene1}}, y = {{gene2}})) + 
    geom_point(alpha = 0.3) +
    facet_wrap(~Type) +
    stat_cor(label.x.npc = .06,
             label.y.npc = 1.0,
             vjust = 1,
             size = 3) +
    geom_smooth(method = "lm")
    
    cor_value |>
      arrange(desc(cor)) |>
      print(n = 3)
    
    print(plot)
}
```
Running this function gives the output below.

```{r function 2 result, message = FALSE, fig.width = 9, fig.height= 8.5, dpi = 300}
scatter_facet(PHGDH, PSAT1)
```