---
title: "TCGA Explorer"
author: "Kyle Dunlap"
date: "2024-02-16"
output: github_document
---

## Introduction

This repository is used to store information about TCGA datasets as well as code on how to mine through them.
In this README, I will share with you several functions that will hopefully help you mine through TCGA data to get some useful results.
These functions use a dataset compiled from the PANCANCER subset in Xenabrowser. If you want to see how I cleaned and put this dataset together, please check out the code in this repository. This pre-cleaned dataset is stored on the University of Utah Biochemistry Fileserver. It is comprised of 12,724 patients, 20,532 genes, and 2 columns at the beginning indicating the TCGA cancer code (https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations) and the abbreviated patient ID. Below is the code used to load in data, and what the first six rows of the table looks like.

```{r load in data, cache = TRUE, message = FALSE}
library(tidyverse)
load("all.RData")

```

```{r make head, echo = FALSE, message = FALSE}
library(tidyverse)
all <- all2
head(all)
```
This dataset is named "all" and will be the main input for all of the following functions.


# Function 1 - scatter() - Makes a scatterplot of two genes.

To use this function, you input two genes and then the cancer type (ACC, BLCA, LIHC) for example. If you want to make a large plot utilizing all 36 cancer types, you will plug in "TCGA" as the third argument.

``` {r function 1, cache = TRUE}
scatter <- function(gene1, gene2, code) {
  
  gene1 <- ensym(gene1)
  gene2 <- ensym(gene2)
  code <- ensym(code)

  filtered_data <- all |>
    select({{gene1}}, {{gene2}}, Type, sample_type) |>
    filter(if (code != "TCGA") Type == code else TRUE) |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic")
  
  cor_value <- cor(filtered_data[[gene1]], filtered_data[[gene2]])
  
  ggplot(filtered_data, aes(x = {{gene1}}, y = {{gene2}})) + 
    geom_point(alpha = 0.2) +
    annotate("text", x = min(filtered_data[[gene1]]), y = max(filtered_data[[gene2]]),
             label = paste(code, "Correlation Coefficient: r =", round(cor_value, 3)),
             hjust = 0, vjust = 0, size = 5.5) +
    geom_smooth(method = "lm") +
    theme_classic() +
    theme(axis.text.x = element_text(size = 15)) +
    theme(axis.text.y = element_text(size = 15)) +
    theme(axis.title.x = element_text(size = 15)) +
    theme(axis.title.y = element_text(size = 15)) 
}
```
Running this function gives the output below.

```{r function 1 result, message = FALSE, out.width = "75%", out.height = "75%", dpi = 300}
scatter(PHGDH, PSAT1, TCGA)
```

# Function 2 - scatter_facet() - Makes scatter plots of 2 genes for all 36 cancer types.
This function takes advantage of the facet_wrap() function in ggplot2. You will input two genes, and get 36 "mini-scatter plots", one for each cancer type. It will return both a table and a plot. To save space, I only show a tiny snippet of the plot, but you can see the whole thing by changing print(n = 3) to print(n = Inf).

```{r function 2}
scatter_facet <- function(gene1, gene2) {
  gene1 <- ensym(gene1)
  gene2 <- ensym(gene2)
  
  
  library(ggpubr)
  filtered_data <- all |>
    select({{gene1}}, {{gene2}}, Type, sample_type) |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic")
  
  cor_value <- filtered_data |>
    group_by(Type) |>
    summarize(
      cor = cor({{gene1}}, {{gene2}})
    )
  
  plot <- ggplot(filtered_data, aes(x = {{gene1}}, y = {{gene2}})) + 
    geom_point(alpha = 0.3) +
    facet_wrap(~Type) +
    stat_cor(label.x.npc = .06,
             label.y.npc = 1.0,
             vjust = 1,
             size = 3) +
    geom_smooth(method = "lm")
    
    cor_value |>
      arrange(desc(cor)) |>
      print(n = 3)
    
    print(plot)
}
```
Running this function gives the output below.

```{r function 2 result, message = FALSE, fig.width = 9, fig.height= 8.5, dpi = 300}
scatter_facet(PHGDH, PSAT1)
```

# Function 3 - single_cor() Gives you the correlation value of two genes.
This function outputs a single number: The correlation value between two genes within one cancer type or across all of them.
```{r function 3}
single_cor <- function(Gene1, Gene2, code) {
  code <- ensym(code)
  all1 <- all |>
    select({{Gene1}}, {{Gene2}}, Type, sample_type) |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic") |>
    filter(if (code != "TCGA") Type == code else TRUE) |>
    summarize(
      corvalue = cor({{Gene1}}, {{Gene2}})
    )
  return(all1)
}
```
Running this code will give you your single value, in this case, the correlation between PSAT1 and PHGDH over all cancer types is 0.476
```{r function 3 result, message = FALSE}
single_cor(PSAT1, PHGDH, TCGA)
```

# Function 4 - boxplot_many() - Give any number of genes and the TCGA type and get a single boxplot 
This can be done pretty easily with XenaBrowser, but this method saves clicking time and improves the graphics. You can input any number of genes (greater than 1) and get a boxplot. As we've been doing, you can input either one cancer type or all of them with "TCGA".

```{r function 4}
boxplot_many <- function(code, ...) {
  genes <- ensyms(...)
  code <- ensym(code)
  genes <- as.character(genes)
  all |>
    select(genes, Type, sample_type) |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic") |>
    filter(if (code != "TCGA") Type == code else TRUE) |>
    pivot_longer(
      cols = (genes),
      names_to = "Gene", 
      values_to = "log2exp"
    ) |>
    mutate(Gene = factor(Gene, levels = genes)) |>
    ggplot(aes(x = Gene, y = log2exp, fill = Gene)) + 
    geom_boxplot() +
    labs(title = code) +
    theme_minimal() +
    theme(axis.line = element_line(linewidth = .5)) +
    theme(axis.text = element_text(size = 12)) +
    theme(legend.position = "none")
}
```
Running this code with any number of genes will give you the boxplot below. The cancer type (or TCGA) must be in the first argument.
``` {r function 4 result, dpi = 300}
boxplot_many(TCGA, OTC, ASS1, SLC7A5, ARG1, ARG2)
```

# Function 5 - boxplot_onegene() -  Give one gene and get a boxplot of all 36 cancer types.
This plot takes one gene as an input and gives you a boxplot of 36 boxes, each corresponding to a cancer type. This is a good way to see what cancers your gene is highly or lowly expressed in.
```{r function 5}
boxplot_onegene <- function(Gene) {
  all |>
    select({{Gene}}, Type, sample_type) |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic") |>
    ggplot(aes(x = reorder(Type, {{Gene}}, FUN=median), y = {{Gene}}, fill = Type)) +
    geom_boxplot() +
    labs(x = "TCGA Code") +
    theme_minimal() +
    theme(axis.line = element_line(linewidth = .5), 
          axis.text.x = element_text(size = 9, angle = 45, hjust = 1, face = "bold"),
          legend.position = "none")
}
```
Here, you input one gene to get the boxplot.
``` {r function 5 result, dpi = 300}
boxplot_onegene(PHGDH)
```

# Function 6 - onegene_corloop() - Correlate over all genes after giving one gene.
This function takes one gene as an input and returns a table of the most correlated gene to the input gene. Like previous functions, you can choose what cancer type to look at. This is a good way to see what genes may have some sort of co-regulation with the input gene.
``` {r function 6}
onegene_corloop <- function(gene, code) {
  code <- ensym(code)
  alltest <- all |>
    filter(sample_type != "Solid Tissue Normal") |>
    filter(sample_type != "Recurrent Tumor") |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic") |>
    select(!20533:20571) |>
    select(!patient) |>
    filter(if (code != "TCGA") Type == code else TRUE) |>
    relocate({{gene}}, .after = Type)
  
  alllist <- lapply(c(2:ncol(alltest)), function(x) cor(alltest[,2], alltest[,x], method = "pearson", use = "complete.obs"))
  listframe <- data.frame(alllist)
  listframe |>
    pivot_longer(
      cols = 1:ncol(listframe),
      names_to = "gene",
      values_to = "cor"
    ) |>
    arrange(desc(cor)) |>
    print(n = 5)
  
}
```
Give the input gene and cancer code. You can print as many genes as you want, but I will only show 5.

```{r function 6 result, message = FALSE, warning = FALSE}
onegene_corloop(PHGDH, TCGA) 
```

# 7. normal_cancer - Create a Boxplot of Solid Tissue vs. Cancer in Your Type
What's cool about this dataset is that there is also tissue next to the diseased area that we have access to. Here, you can plot the values of a gene in a cancer tissue vs. the healthy tissue nearby.

```{r function 7}
normal_cancer <- function(Gene, code){
  code <- ensym(code)
  title1 <- paste0("Comparison of ", code, " to nearby Non-Cancerous Tissue")
  all |>
    filter(sample_type != "Additional - New Primary") |>
    filter(sample_type != "Additional Metastatic") |>
    filter(sample_type != "Recurrent Tumor") |>
    select({{Gene}}, Type, sample_type) |>
    filter(if (code != "TCGA") Type == code else TRUE) |>
    ggplot(aes(x = sample_type, y = {{Gene}}, fill = sample_type)) + 
    geom_boxplot() +
    labs(title = title1) +
    theme_minimal() +
    theme(axis.line = element_line(linewidth = .5)) +
    theme(axis.text = element_text(size = 12)) +
    theme(axis.text.x = element_text(size = 8.5)) +
    theme(legend.position = "none")
}
```

Give a gene and a cancer type to get a boxplot.
```{r function 7 result}
normal_cancer(PHGDH, BRCA)
```

# 8. rank - Ranks cancer types by expression of given gene
This function will take one gene and rank its expression within the 36 available cancer types in TCGA.
```{r function 8}
rank <- function(gene) {
  gene <- ensym(gene)
  all |>
  select({{gene}}, Type) |>
  group_by(Type) |>
  summarize(
    mean = mean({{gene}}, na.rm = TRUE) 
  ) |>
  arrange(desc(mean)) |>
  print(n = 5)
}
```
Input one gene. You will only see the top 5 results.
```{r function 8 result}
rank(PHGDH)
```